Multi-state modelling with msm: a practical course

### 2021-09-23 docs/msmcourse_src/02-msm-basic.Rmd
More course edits

* occupied state 2 throughout the period between years 2 and 3.09 and between years 4 and 5 (`obstype` 2)

@@

A risk of the kinds of models used here is that they may be weakly identifiable from the data.  Often the optimisation will not converge, or converge to an answer that its not sensible.

@@

In this case the model should be simplified, e.g. by combining states, removing covariates, constraining intensities to be equal to each other (`qconstraint` option to `msm`) or constraining covariate effects (see the [chapter on covariates](#covariates))

>>>>

* occupied state 2 throughout the period between years 2 and 3.09

* occupied state 2 between years 4 and 5 (`obstype` 2) before moving to state 3 at time 5

@@

A risk of the kinds of models used here is that they may be weakly identifiable from the data.  Often the optimisation will not converge, or converge to an answer that is not useful.

@@

If it doesn't converge, that usually means it is sensible to simplify the model, e.g. by combining states, removing covariates, constraining intensities to be equal to each other (`qconstraint` option to `msm`) or constraining covariate effects (see the [chapter on covariates](#covariates))

### 2021-09-23 docs/msmcourse_src/03-outputs.Rmd
More course edits

    Repeatedly draw from the asymptotic sampling distribution of the estimates, which is a multivariate normal with mean defined by the maximum likelihood estimates, and covariance defined by the (inverse) Hessian at the maximised log likelihood.

@@

* ? years in state 1

* ? years in state 2 ...

@@

    Note also that in this example, some aspects of the goodness-of-fit checks are misleading - see the solution for an explanation why.

@@

1.  Picking a range of prediction times covering those in the data (1 year, 5 years, 10 years), we can compute the probability of death (state 4 in the model) by this time by extracting the fourth column of the transition probability matrix.

@@

    In fact, this is an artefactual finding resulting from people in the data being followed up for different lengths of time.  To calculate the probability at each time, `msm` uses a denominator defined by the number of people whose state we know at this time.   This overestimates the probability of death, because the people whose state we know at later times are more likely to be the people who have died.  We don't know the state at later times for people whose outcomes were censored at the end of the study when they were still alive.

>>>>

    Repeatedly draw from the asymptotic sampling distribution of the estimated log intensities (and covariate effects), which is a multivariate normal with mean defined by the maximum likelihood estimates, and covariance defined by the (inverse) Hessian at the maximised log likelihood.

@@

* ? years in state $s=1$

* ? years in state $s=2$ ...

@@

@@

1.  Picking a range of prediction times covering those in the data (e.g. 1 year, 5 years, 10 years), we can compute the probability of death (state 4 in the model) by this time by extracting the fourth column of the transition probability matrix.

@@

    In fact (though you were not expected to know this from the information given in the question!) this is an artefactual finding resulting from people in the data being followed up for different lengths of time.  To calculate the probability at each time, `msm` uses a denominator defined by the number of people whose state we know at this time.   This overestimates the probability of death, because the people whose state we know at later times are more likely to be the people who have died.  We don't know the state at later times for people whose outcomes were censored at the end of the study when they were still alive.

### 2021-09-23 docs/msmcourse_src/04-covariates.Rmd
More course edits

Most general model for the psoriatic arthritis data

```{tikz, fig.cap = "Psoriatic arthritis", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\definecolor{deepskyblue}{rgb}{0, 0.75, 1}
\tikzstyle{state}=[minimum size = 0.8cm, draw, rounded corners, fill=deepskyblue]
\begin{tikzpicture}[]
\node [state] (state1) {1. 0-1 damaged joints};
\node [state, right=of state1] (state2) {2. 1-4 damaged joints};
\node [state, right=of state2] (state3) {3. 5-9 damaged joints};
\node [state, right=of state3] (state4) {4. 10+ damaged joints};
\draw[->] ($(state1.east)+(0cm,0.1cm)$) -- ($(state2.west)+(0cm,0.1cm)$) node [midway,above=2em] {$q_{12}^{(0)} \exp(\boldsymbol{\beta}_{12}^{\prime} \mathbf{z})$};
\draw[->] ($(state2.east)+(0cm,0.1cm)$) -- ($(state3.west)+(0cm,0.1cm)$) node [midway,above=2em] {$q_{23}^{(0)} \exp(\boldsymbol{\beta}_{23}^{\prime} \mathbf{z})$};
\draw[->] ($(state3.east)-(0cm,0.1cm)$) -- ($(state4.west)-(0cm,0.1cm)$) node [midway,above=2em] {$q_{34}^{(0)} \exp(\boldsymbol{\beta}_{34}^{\prime} \mathbf{z})$}; 
\end{tikzpicture}
```

Consider which transitions may be affected by which covariates.  Don't just put all covariates on all transitions without thinking --- you may not have enough data informing transitions out of particular states.

@@

Given the AICs and the estimates from the full model, it seems sufficient to assume that `hieffusn` is a predictor of the 1-2 and 2-3 transitions, and `ollwsdrt` is a predictor of the 2-3 transition.

@@

1. First we start with the full model where the covariate affects all transitions, and examine the hazard ratios under this model.

@@

    We see that the pre-transplant IHD is associated with an increased risk of CAV onset (state 1 - state 2) and death before CAV onset (state 1 - state 4).    The confidence intervals for the hazard ratios of IHD on the other transition rates are wide and include the null 1.

@@

    The passage probability to severe CAV (which we computed in the previous chapter) is similar between both groups of people (`IHD=0`) and (`IHD=1`), though people who have had IHD are estimated to be slightly more likely to go through the "severe CAV" state before they die.

>>>>

There are three transitions in the model.  Which transitions should be affected by which covariates?

```{tikz, fig.cap = "Psoriatic arthritis", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\definecolor{deepskyblue}{rgb}{0, 0.75, 1}
\tikzstyle{state}=[minimum size = 0.8cm, draw, rounded corners, fill=deepskyblue]
\begin{tikzpicture}[]
\node [state] (state1) {1. 0-1 damaged joints};
\node [state, right=of state1] (state2) {2. 1-4 damaged joints};
\node [state, right=of state2] (state3) {3. 5-9 damaged joints};
\node [state, right=of state3] (state4) {4. 10+ damaged joints};
\draw[->] ($(state1.east)+(0cm,0.1cm)$) -- ($(state2.west)+(0cm,0.1cm)$) node [midway,above=2em] {$q_{12}^{(0)} \exp(\boldsymbol{\beta}_{12}^{\prime} \mathbf{z})$};
\draw[->] ($(state2.east)+(0cm,0.1cm)$) -- ($(state3.west)+(0cm,0.1cm)$) node [midway,above=2em] {$q_{23}^{(0)} \exp(\boldsymbol{\beta}_{23}^{\prime} \mathbf{z})$};
\draw[->] ($(state3.east)+(0cm,0.1cm)$) -- ($(state4.west)+(0cm,0.1cm)$) node [midway,above=2em] {$q_{34}^{(0)} \exp(\boldsymbol{\beta}_{34}^{\prime} \mathbf{z})$}; 
\end{tikzpicture}
```

Don't just put all covariates on all transitions without thinking --- you may not have enough data informing transitions out of particular states.

@@

It seems sufficient to assume that `hieffusn` is a predictor of the 1-2 and 2-3 transitions, and `ollwsdrt` is a predictor of the 2-3 transition

Check the coefficients from the fitted models to verify that this is plausible...

@@

1. We convert the categorical variable `pdiag` to a binary 0/1 covariate indicating primary diagnosis of IHD.   First we fit the full model where the binary covariate affects all transitions, and examine the hazard ratios under this model.

@@

    We see that the pre-transplant IHD is associated with an increased risk of CAV onset (state 1 - state 2), and possibly also with death before CAV onset (state 1 - state 4).    The confidence intervals for the hazard ratios of IHD on the other transition rates are wide and include the null 1.

@@

    The passage probability to severe CAV (which we computed in the previous chapter), i.e. the chance of going through the "severe CAV" state before death, is similar between both groups of people (`IHD=0`) and (`IHD=1`), though slightly higher for people who have had IHD.

### 2021-09-23 docs/msmcourse_src/05-timedep.Rmd
More course edits

 - times when the age changes (hence when the intensity changes) (`times`)

 - the values that age takes at these times (`covariates`)

@@

A finer time grid will give a better approximation to a prediction from the ideal model where risk changes _continuously_ with age.

@@

@@

Problem is that the state is unknown at 6 and 12 years

@@

Replace the `NA` values for the unknown state with a code (99, say), and use the following syntax to tell `msm` what states this code can represent:

@@

* (as here) time-dependent covariates that change at a common time.  We know the covariate value, but not the state, at that time.

Footnote: syntax uses the term `censor`, because a _censored_ random variable is one that is partially observed, but in `msm` this refers to a partially observed state, not a partially-known event time.

## Shorthand for defining this model in `msm`

@@

No evidence for time dependence in the intensities

@@

Here, when we predict the transition probabilities over 15 years, it will automatically adjust for the intensities changing at 5 and 10 years.   So there is no need to use `pmatrix.piecewise.msm` or supply a `covariates` argument.

@@

## Variants of pci models

What if we want to build more detailed models with time-dependence, e.g.

* models with a different covariate effect for each time period?

* models where the time period affects some transition rates but not others?

1. Construct the expanded dataset

    Fit a standard `pci` model and then use the `model.frame` function to extract the expanded dataset constructed by `msm`, with

    - extra rows at the time change points, and

@@

2. Fit model to the expanded dataset, using `censor`

@@

    - here the code is 5, so call `msm` with `censor = 5`.

@@

Note if you get a `numerical overflow` error, this may be solved by normalising the optimisation using `msm(..., control=list(fnscale=4000))`, as described at the end of Chapter 2, noting that the -2 log likelihood in this example takes a value of around 4000.

>>>>

 - times when the `age` covariate changes (hence when the intensity changes) (`times`)

 - the values that `age` takes at these times (`covariates`)

@@

`times` doesn't have to be an integer.  A finer time grid might be chosen, to give a better approximation to a prediction from the ideal model where risk changes _continuously_ with age.

@@

Then we could predict future outcomes for someone with a particular combination of states and risk factors (= someone in a particular state in the expanded model).

@@

Problem is that the state is unknown at 6 and 12 years if the data are intermittently observed.

@@

Replace the `NA` values in the data for the unknown state with a code (99, say), and use the following syntax to fit the model while telling `msm` what states this code can represent:

@@

* (as in the example here) time-dependent intensities that change at a common time.  We know the value of the "time period" covariate, but not the state, at that time.

(Footnote: `msm` syntax uses the term `censor`, because a _censored_ random variable is one that is partially observed, but in `msm` this refers to a partially observed state, not a partially-known event time.)

## `pci`: piecewise constant intensities (which change at common times)

`msm` provides a shorthand for these models where the intensities depend on the time period.

@@

No evidence for time dependence in the intensities in this example

@@

Here, when we predict the transition probabilities over 15 years, it will automatically adjust for the intensities changing at 5 and 10 years.

So there is no need to use `pmatrix.piecewise.msm` or supply a `covariates` argument.

@@

## Variants of `pci` models (advanced)

What if we want to build more detailed models with time-dependent intensities, e.g.

* models with a second covariate, whose effect is different within each time period?  i.e. a time/covariate interaction.

* models where the time period affects some transition intensities but not others?

1. **Construct the expanded dataset**

    Fit a standard `pci` model, then use `model.frame()` to extract the expanded dataset constructed by `msm`, that has

    - extra rows at the time change points where the state is unknown,

@@

2. **Fit model to the expanded dataset**, using `censor`

@@

    - here the "state unknown" code is 5, so call `msm` with `censor = 5`.

@@

Note if you get a `numerical overflow` error, this may be solved by normalising the optimisation using `msm(..., control=list(fnscale=4000))`, as described at the end of Chapter 2, noting that the -2 log-likelihood being optimised in this example takes a value of around 4000.

### 2021-09-23 docs/msmcourse_src/06-advanced.Rmd
More course edits

```{tikz, fig.cap = "Hidden Markov model", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\definecolor{deepskyblue}{rgb}{0, 0.75, 1}
\tikzstyle{state}=[minimum size = 0.8cm, draw, rounded corners, fill=deepskyblue]
\tikzstyle{obsstate}=[state]
\tikzstyle{hidstate}=[state, fill=gray!20]
\tikzstyle{miscarrow}=[dotted,draw=gray,line width=1mm]
\begin{tikzpicture}[]
\node [hidstate] (well) {1. No disease};
\node [hidstate, right=of well] (mild) {2. Mild disease};
\node [hidstate, right=of mild] (severe) {3. Severe disease};
\node [obsstate, below=of mild] (death) {4. Death};
\draw[->] (well) -- (mild) node [] {};
\draw[->] (mild) -- (severe) node [] {};
\node [obsstate, above=of well] (obswell) {Observed ``well''};
\node [obsstate, above=of mild] (obsmild) {Observed ``mild''};
\node [obsstate, above=of severe] (obssev) {Observed ``severe''};
\draw[miscarrow] (well) -- (obswell);
\draw[miscarrow] (well) -- (obsmild);
\draw[miscarrow] (mild) -- (obswell);
\draw[miscarrow] (mild) -- (obsmild);
\draw[miscarrow] (mild) -- (obssev);
\draw[miscarrow] (severe) -- (obsmild);
\draw[miscarrow] (severe) -- (obssev);
\draw[->] (well) -- (death) node [midway, left] {};
\draw[->] (mild) -- (death) node [midway] {};
\draw[->] (severe) -- (death) node [midway, right] {};
\end{tikzpicture}
```

@@

To relax the Markov assumption for a specific state, split up that state into hidden "phases" and define a Markov model on a state space defined by the phases and the remaining states.  Transition time between the phases is unknown.

@@

This is a (partially) hidden Markov model, hence can be implemented in `msm`.

@@

- Several R packages available for random effects models with _time-to-event_ data - see the [CRAN task view](https://cran.r-project.org/web/views/Survival.html).

>>>>

```{tikz, fig.cap = "Hidden Markov multi-state model where states 1,2,3 can be misclassified", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\definecolor{deepskyblue}{rgb}{0, 0.75, 1}
\tikzstyle{state}=[minimum size = 0.8cm, draw, rounded corners, fill=deepskyblue]
\tikzstyle{obsstate}=[state]
\tikzstyle{hidstate}=[state, fill=gray!20]
\tikzstyle{miscarrow}=[dotted,draw=gray,line width=1mm]
\begin{tikzpicture}[]
\node [hidstate] (well) {1. No disease};
\node [hidstate, right=of well] (mild) {2. Mild disease};
\node [hidstate, right=of mild] (severe) {3. Severe disease};
\node [obsstate, below=of mild] (death) {4. Death};
\draw[->] (well) -- (mild) node [] {};
\draw[->] (mild) -- (severe) node [] {};
\node [obsstate, above=of well] (obswell) {Observed ``well''};
\node [obsstate, above=of mild] (obsmild) {Observed ``mild''};
\node [obsstate, above=of severe] (obssev) {Observed ``severe''};
\draw[miscarrow] (well) -- (obswell);
\draw[miscarrow] (well) -- (obsmild);
\draw[miscarrow] (mild) -- (obswell);
\draw[miscarrow] (mild) -- (obsmild);
\draw[miscarrow] (mild) -- (obssev);
\draw[miscarrow] (severe) -- (obsmild);
\draw[miscarrow] (severe) -- (obssev);
\draw[->] (well) -- (death) node [midway, left] {};
\draw[->] (mild) -- (death) node [midway] {};
\draw[->] (severe) -- (death) node [midway, right] {};
\end{tikzpicture}
```

@@

To relax the Markov assumption for a specific state, split up that state into hidden "phases" and define a Markov model on a state space defined by the phases and the remaining states.  Transition time between the phases is unknown. 
For example:

@@

This is equivalent to making the sojourn distribution in state 2 more flexible than an exponential distribution, so the hazard of leaving state 2 depends on how long has been spent there.

It is a (partially) hidden Markov model, hence can be implemented in `msm` using its hidden Markov model machinery.

@@

- Several R packages available for random effects models with _time-to-event_ data - e.g. `frailtypack` and others in the [CRAN task view](https://cran.r-project.org/web/views/Survival.html).
